Modern type-safe embedding of attribute grammars

### 2018-12-21 doc/Paper.tex
Update PDF to 65e6608afa4f794444349fe633d97c4bba31428a

\usepackage{amsmath}
\usepackage{calc}
\usepackage{csquotes}
\usepackage{color}
\usepackage{comment}
\usepackage{float}
\usepackage{graphicx}
\usepackage{url}
\usepackage{subcaption}
\usepackage{wrapfig}
\usepackage{pbox}

@@

@@

>>>>

\usepackage{amsmath}
\usepackage{calc}
\usepackage{csquotes}
\usepackage{color}
\usepackage{comment}
\usepackage[inline,shortlabels]{enumitem}
\usepackage{float}
\usepackage{graphicx}
\usepackage{url}
\usepackage{subcaption}
\usepackage{wrapfig}
\usepackage{pbox}

@@

\newcommand{\Let}{{\sf LET}}

@@

\section{Running example}

\newcommand{\withSub}[2]{{\parbox{\widthof{#1#2}}{#1$_\text{#2}$}}}
\newcommand{\itWithSub}[2]{{\parbox{\widthof{#1#2}}{\textit{#1}$_\text{#2}$}}}

In order to illustrate the convenience of our embedding, we will consider the
problem of analyzing the semantic correctness of certain functional programming
excerpts. In particular, we will focus on \text{\texfamily {\bfseries let}} expressions that are common in
functional languages like Haskell~\cite{}, ML~\cite{}, or Scala~\cite{}.
Throughout the paper, we shall refer to these expressions as programs in the
\Let\ language.

A program in the \Let\ language consists of instruction blocks, where each
instruction either:
\begin{enumerate*}[i)]
  \item declares a variable or
  \item defines a nested block.
\end{enumerate*}
When declaring a variable, the programmer must also assigned a value to it; such
value may be a constant, the value of a variable or an expression on these
elements. Ultimately, each block defines a value as an expression on variables
defined in it.

Examples of programs in \Let\ are given next.

\begin{tabbing}\texfamily
\withSub{program}{1}~=~{\bfseries let}~x~=~4~{\bfseries in}~x\\
\texfamily \\
\texfamily \withSub{program}{2}~=~{\bfseries let}~x~=~4~{\bfseries in}~x~+~3\\
\texfamily \\
\texfamily \withSub{program}{3}~=\\
\texfamily ~~{\bfseries let}~x~=~4\\
\texfamily ~~~~~~y~=~{\bfseries let}~w~=~2~{\bfseries in}~x~+~w\\
\texfamily ~~{\bfseries in}~~x~+~y
\end{tabbing}

The values associated with \text{\texfamily \withSub{program}{1}} and \text{\texfamily \withSub{program}{2}} are, quite
straightforwardly, \text{\texfamily 4} and \text{\texfamily 7}, respectively. As of \text{\texfamily \withSub{program}{3}}, its value of
\text{\texfamily 10} is calculated adding the value of \text{\texfamily x}, which is \text{\texfamily 4} and the value of \text{\texfamily y},
i.e., \text{\texfamily 6}. The value of \text{\texfamily y} is obtained by adding the value of \text{\texfamily w}, which is
\text{\texfamily 2}, and the value of \text{\texfamily x}, which, again, is \text{\texfamily 4}.

Our goal is to implement a semantic analyzer that deals with the scope rules of
the \Let\ language. These rules are quite natural and can be described as:
\begin{enumerate}
  \item if a variable is used in a block, it must be declared in that same block
  or in an outer one. The declaration of a variable, however, may occur after its
  use;
  \item a variable identifier may be declared at most once within the same block.
  In an inner block, declaring a variable that has already been declared in an
  outer one is allowed: the identifier in the local scope hides the definition of
  the same identifier in the global one.
\end{enumerate}

Let us now consider a more complex \Let\ program:

\begin{tabbing}\texfamily
\withSub{program}{e}~=\\
\texfamily ~~{\bfseries let}~x~=~y\\
\texfamily ~~~~~~a~=~{\bfseries let}~y~=~4~{\bfseries in}~y~+~w\\
\texfamily ~~~~~~x~=~5\\
\texfamily ~~~~~~y~=~6\\
\texfamily ~~{\bfseries in}~~x~+~a
\end{tabbing}

According to the scope rules that we have just defined, \text{\texfamily \withSub{program}{e}} contains two
errors:
\begin{enumerate*}[1)]
  \item at the outer block, variable \text{\texfamily x} has been declared twice and
  \item at the inner block, the use of variable \text{\texfamily w} has no binding occurrence at
        all.
\end{enumerate*}
Notice that \text{\texfamily y} has been declared at both the inner and the outer levels, which
in itself is not a problem (the inner declaration hides the outer one).

Programs such as the ones we have presented describe the basic block-structure
found in many languages, with the peculiarity that variables can be used before
they are defined.

We aim to implement a program that analyses \Let\ programs and computes a list
containing the identifiers which do not obey the scope rules of the language. In
order to facilitate the debugging phase, we require that the list of invalid
identifiers follows the sequential structure of the program. That is to say,
e.g., that the semantic meaning of processing \text{\texfamily \withSub{program}{e}} is \text{\texfamily [w,\Sp x]}: the use
of the undeclared variable \text{\texfamily w} occurs in line 3 whereas the duplicate
declaration of \text{\texfamily x} occurs in line 4.

Since variables can be used before they are declared, a natural way to implement
such an analysis is to traverse a \Let\ program twice: once to accumulate the
declarations of identifiers (at each block), and again to check the uses of
identifiers against such declarations. The uniqueness of declarations can
already be detected in the first traversal: for each newly encountered
declaration, it is possible to check whether its identifier has already been
declared (in the same level).

In order to implement such program, we first need a representation for programs
in the \Let\ language. For this, we may use the following Haskell data-types:

\begin{tabbing}\texfamily
{\bfseries type}~{\itshape Var}~~~~=~{\itshape String}\\
\texfamily {\bfseries data}~{\itshape Let}~~~~=~{\itshape Let}~{\itshape Decls}~{\itshape Expr}\\
\texfamily {\bfseries data}~{\itshape Decls}~~=~~{\itshape Empty}\\
\texfamily ~~~~~~~~~~~~|~~{\itshape Cons}~~~~{\itshape Var}~{\itshape Expr}~~{\itshape Decls}\\
\texfamily ~~~~~~~~~~~~|~~{\itshape Nested}~~{\itshape Var}~{\itshape Let}~~~{\itshape Decls}\\
\texfamily {\bfseries data}~{\itshape Expr}~~~=~~{\itshape Const}~~~~~{\itshape Int}\\
\texfamily ~~~~~~~~~~~~|~~{\itshape Variable}~~{\itshape Var}\\
\texfamily ~~~~~~~~~~~~|~~{\itshape Plus}~~~{\itshape Expr}~{\itshape Expr}\\
\texfamily ~~~~~~~~~~~~|~~{\itshape Times}~~{\itshape Expr}~{\itshape Expr}
\end{tabbing}

In this representation, \text{\texfamily \withSub{program}{e}} above is defined as:

\begin{tabbing}\texfamily
\withSub{program}{e}~=~{\itshape Let}\\
\texfamily ~~({\itshape Cons}\\
\texfamily ~~~~\char34 x\char34 \\
\texfamily ~~~~({\itshape Variable}~\char34 y\char34 )\\
\texfamily ~~~~({\itshape Nested}\\
\texfamily ~~~~~~\char34 a\char34 \\
\texfamily ~~~~~~({\itshape Let}~({\itshape Cons}~\char34 y\char34 ~({\itshape Const}~4)~{\itshape Empty})~({\itshape Plus}~({\itshape Variable}~\char34 y\char34 )~({\itshape Variable}~\char34 w\char34 )))\\
\texfamily ~~~~~~({\itshape Cons}~\char34 x\char34 ~({\itshape Const}~5)~({\itshape Cons}~\char34 y\char34 ~({\itshape Const}~6)~{\itshape Empty}))\\
\texfamily ~~~~)\\
\texfamily ~~)\\
\texfamily ~~({\itshape Plus}~({\itshape Variable}~\char34 x\char34 )~({\itshape Variable}~\char34 a\char34 ))
\end{tabbing}

Now, we implement name analysis on an abstract \Let\ tree using a set of
composable functions.

The function that implements the first traversal described above needs to pass
all the information that is required for the second traversal. Namely, in order
to compute the final list of errors in the desired order, the second traversal
needs to \textit{where} errors have occurred during the first. Also, the value
of the nesting level must also be carried around because it is on the second
traversal that the first traversal on nested expressions starts (the initial
environment of an inner block is composed by the complete environment of its
outer one).

In order to make available all the information that the second traversal needs,
the first traversal will build a structure such as:

\begin{tabbing}\texfamily
{\bfseries data}~\itWithSub{Let}{2}~~~=~\itWithSub{Let}{2}~\itWithSub{Decls}{2}~{\itshape Expr}\\
\texfamily {\bfseries data}~\itWithSub{Decls}{2}~=~\itWithSub{Empty}{2}\\
\texfamily ~~~~~~~~~~~~|~\itWithSub{Cons}{2}~{\itshape Errors}~{\itshape Expr}~\itWithSub{Decls}{2}\\
\texfamily ~~~~~~~~~~~~|~\itWithSub{Nested}{2}~{\itshape Errors}~{\itshape Lev}~{\itshape Let}~\itWithSub{Decls}{2}
\end{tabbing}

We are now ready to implement the two traversal strategy that we have described.

\begin{tabbing}\texfamily
{\bfseries type}~{\itshape Errors}~~=~[{\itshape String}]\\
\texfamily {\bfseries type}~{\itshape Lev}~~~~~=~{\itshape Int}\\
\texfamily \\
\texfamily semantics~::~{\itshape Let}~\WithMath{->}{\rightarrow}~{\itshape Errors}\\
\texfamily semantics~program~=~errors\\
\texfamily ~{\bfseries where}\\
\texfamily ~~(\withSub{let}{2},~env)~=~\withSub{duplicate}{\textit{Let}}~program~[]~0\\
\texfamily ~~errors~~~~~~~=~\withSub{missing}{\textit{Let}}~\withSub{let}{2}~env\\
\texfamily \\
\texfamily \withSub{duplicate}{\textit{Let}}~::~{\itshape Let}~\WithMath{->}{\rightarrow}~[({\itshape Var},~{\itshape Lev})]~\WithMath{->}{\rightarrow}~{\itshape Lev}~\WithMath{->}{\rightarrow}~(\itWithSub{Let}{2},~[({\itshape Var},~{\itshape Lev})])\\
\texfamily \withSub{duplicate}{\textit{Let}}~({\itshape Let}~decls~expr)~dcli~lev~=~(\itWithSub{Let}{2}~\withSub{decls}{2}~expr,~dclo)\\
\texfamily ~~{\bfseries where}~(\withSub{decls}{2},~dclo)~=~\withSub{duplicate}{\textit{Decls}}~decls~dcli~lev\\
\texfamily \\
\texfamily \withSub{duplicate}{\textit{Decls}}~::~{\itshape Decls}~\WithMath{->}{\rightarrow}~[({\itshape Var},~{\itshape Lev})]~\WithMath{->}{\rightarrow}~{\itshape Lev}~\WithMath{->}{\rightarrow}~(\itWithSub{Decls}{2},~[({\itshape Var},~{\itshape Lev})])\\
\texfamily \withSub{duplicate}{\textit{Decls}}~{\itshape Empty}~dcli~lev~=~(\itWithSub{Empty}{2},~dcli)\\
\texfamily \withSub{duplicate}{\textit{Decls}}~({\itshape Cons}~var~expr~decls)~dcli~lev~=\\
\texfamily ~~(\itWithSub{Cons}{2}~\textit{error}~expr~\withSub{decls}{2},~dclo)\\
\texfamily ~{\bfseries where}\\
\texfamily ~~\textit{error}~~~~~~~~~~~=~{\bfseries if}~(var,~lev)~`elem`~dcli~{\bfseries then}~[var]~{\bfseries else}~[]\\
\texfamily ~~(\withSub{decls}{2},~dclo)~=~\withSub{duplicate}{\textit{Decls}}~decls~((var,~lev)~:~dcli)~lev\\
\texfamily \withSub{duplicate}{\textit{Decls}}~({\itshape Nested}~var~nested~decls)~dcli~lev~=\\
\texfamily ~~(\itWithSub{Nested}{2}~\textit{error}~(lev~+~1)~nested~\withSub{decls}{2},~dclo)\\
\texfamily ~{\bfseries where}\\
\texfamily ~~\textit{error}~~~~~~~~~~~=~{\bfseries if}~(var,~lev)~`elem`~dcli~{\bfseries then}~[var]~{\bfseries else}~[]\\
\texfamily ~~(\withSub{decls}{2},~dclo)~=~\withSub{duplicate}{\textit{Decls}}~decls~((var,~lev)~:~dcli)~lev\\
\texfamily \\
\texfamily \withSub{missing}{\textit{Let}}~::~\itWithSub{Let}{2}~\WithMath{->}{\rightarrow}~[({\itshape Var},~{\itshape Lev})]~\WithMath{->}{\rightarrow}~{\itshape Errors}\\
\texfamily \withSub{missing}{\textit{Let}}~(\itWithSub{Let}{2}~decls~expr)~env~=~\withSub{errors}{1}~\WithMath{++}{+\!\!+}~\withSub{errors}{2}\\
\texfamily ~{\bfseries where}\\
\texfamily ~~\withSub{errors}{1}~=~\withSub{missing}{\textit{Decls}}~decls~env\\
\texfamily ~~\withSub{errors}{2}~=~\withSub{missing}{\textit{Expr}}~expr~env\\
\texfamily \\
\texfamily \withSub{missing}{\textit{Decls}}~::~\itWithSub{Decls}{2}~\WithMath{->}{\rightarrow}~[({\itshape Var},~{\itshape Lev})]~\WithMath{->}{\rightarrow}~{\itshape Errors}\\
\texfamily \withSub{missing}{\textit{Decls}}~(\itWithSub{Cons}{2}~\textit{error}~expr~decls)~env~=~\textit{error}~\WithMath{++}{+\!\!+}~errors\\
\texfamily ~~{\bfseries where}~errors~=~\withSub{missing}{\textit{Expr}}~expr~env~\WithMath{++}{+\!\!+}~\withSub{missing}{\textit{Decls}}~decls~env\\
\texfamily \withSub{missing}{\textit{Decls}}~(\itWithSub{Nested}{2}~\textit{error}~lev~nested~decls)~env~=~\textit{error}~\WithMath{++}{+\!\!+}~errors\\
\texfamily ~{\bfseries where}\\
\texfamily ~~(\withSub{nested}{2},~dclo)~=~\withSub{duplicate}{\textit{Let}}~nested~env~lev\\
\texfamily ~~errors~~~~~~~~~~~=~\withSub{missing}{\textit{Let}}~\withSub{nested}{2}~dclo~\WithMath{++}{+\!\!+}~\withSub{missing}{\textit{Decls}}~decls~env\\
\texfamily \withSub{missing}{\textit{Decls}}~\itWithSub{Empty}{2}~\char95 ~=~[]\\
\texfamily \\
\texfamily \withSub{missing}{\textit{Expr}}~::~{\itshape Expr}~\WithMath{->}{\rightarrow}~[({\itshape Var},~b)]~\WithMath{->}{\rightarrow}~{\itshape Errors}\\
\texfamily \withSub{missing}{\textit{Expr}}~({\itshape Const}~\char95 )~\char95 ~=~[]\\
\texfamily \withSub{missing}{\textit{Expr}}~({\itshape Plus}~\withSub{expr}{1}~\withSub{expr}{2})~env~=\\
\texfamily ~~\withSub{missing}{\textit{Expr}}~\withSub{expr}{1}~env~\WithMath{++}{+\!\!+}~\withSub{missing}{\textit{Expr}}~\withSub{expr}{2}~env\\
\texfamily \withSub{missing}{\textit{Expr}}~({\itshape Times}~\withSub{expr}{1}~\withSub{expr}{2})~env~=\\
\texfamily ~~\withSub{missing}{\textit{Expr}}~\withSub{expr}{1}~env~\WithMath{++}{+\!\!+}~\withSub{missing}{\textit{Expr}}~\withSub{expr}{2}~env\\
\texfamily \withSub{missing}{\textit{Expr}}~({\itshape Variable}~var)~env~=~{\bfseries if}~var~`elem`~map~fst~env~{\bfseries then}~[]~{\bfseries else}~[var]
\end{tabbing}

Notice that \text{\texfamily \withSub{duplicate}{\textit{Let}}} not only computes the total environment (using an
initially empty accumulating parameter), but it also computes a \text{\texfamily \itWithSub{Let}{2}}
intermediate data structure that stores, e.g., the duplicated variables detected
during the first traversal. The second traversal starts with a call to
\text{\texfamily \withSub{missing}{\textit{Let}}} giving that computed data structure and the accumulated environment
as arguments. It produces the list of errors that follows the sequential
structure of the program.

In function \text{\texfamily \withSub{duplicate}{\textit{Decls}}}, for every block we compute: its environment, its
level and its invalid identifiers. The environment defines the context where the
block occurs. It consists of all the identifiers that are visible in the block
(annotated with the level of the block). The level indicates the nesting depth
of a block. Observe that we have to distinguish between the same identifier
declared at different levels, which is valid.

Finally, please note that in the second traversal of a nested expression, in
function \text{\texfamily \withSub{missing}{\textit{Decls}}} for the constructor \text{\texfamily \itWithSub{Nested}{2}}, the program performs the
two traversals to the body of that expression: calls \text{\texfamily \withSub{duplicate}{\textit{Let}}} and
\text{\texfamily \withSub{missing}{\textit{Let}}}.

Although the semantic analysis we have implemented for the \Let\ language is
relatively simple, still we had to face some challenges.

Indeed, scheduling computations was by no means trivial, with intermingled
recursive functions. Also, we had to carefully design and implement intermediate
data structures in order to convey data between traversals. These challenges are
common to functional programming solutions to realistic programming problems.

In lazy (functional) programming languages, one can avoid both the need for
scheduling and for additional data structures by constructing circular
programs~\cite{Bird84}. This strategy, however, compromises the much desired
modular nature of the implementations.

In our work, we seek an elegant and efficient alternative to the construction of
functional programs.
